Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-04-04 manuscript.qmd
improve midline ext. formalism

import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from matplotlib import ticker

from lymph import models
from lyscripts import utils

from scripts import shared, paths, COL, CONTRA_LNLS
```

@@

$$
\begin{aligned}
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) &= \sum_{t=0}^{t_\text{max}} p(t) \cdot P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid t \right) \\
&= \sum_{t=0}^{t_\text{max}} p(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
\end{aligned}
$$

@@

$$
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} p(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
$$

@@

To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable. It starts lateralized and every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.

Technically, the introduction of this additional random variable doubles the space of the hidden states. However, since we assume no correlation between the tumor's lateralization and the metastases in the LNLs, we can evolve the two parts separately.

As with the state of the lymphatic network, we can also define a small transition matrix $\mathbf{A}_\epsilon$ that evolves the lateralized starting state $\boldsymbol{\epsilon}[0] = \begin{pmatrix} 1 & 0 \end{pmatrix}$:

$$
\boldsymbol{\epsilon}[t] = \boldsymbol{\epsilon}[0] \cdot \mathbf{A}_\epsilon^t = \begin{pmatrix} 1 & 0 \end{pmatrix} \cdot \begin{pmatrix} 1 - \epsilon & \epsilon \\ 0 & 1 \end{pmatrix}^t = \begin{pmatrix} (1 - \epsilon)^t & 1 - (1 - \epsilon)^t \end{pmatrix}
$$

We can again collect the entire evolution in a matrix $\boldsymbol{\Lambda}_\epsilon$ that stores the probabilies $P(\boldsymbol{\epsilon} \mid t)$ for each time-step $t$:

$$
\boldsymbol{\Lambda}_\epsilon = P(\boldsymbol{\epsilon} \mid t) = \begin{pmatrix}
\boldsymbol{\epsilon}[0] \cdot \mathbf{A}_\epsilon^0 \\
\boldsymbol{\epsilon}[1] \cdot \mathbf{A}_\epsilon^1 \\
\vdots \\
\boldsymbol{\epsilon}[t] \cdot \mathbf{A}_\epsilon^t \\
\end{pmatrix}
$$

This quantity has $t$ rows, one for each time-step, and two columns: one for the decreasing probability of a lateralized tumor and one for the increasing probability of a tumor that has crossed the mid-sagittal line.

```{python}
#| echo: false
#| label: fig-midext-evo
#| fig-cap: The matrix $\boldsymbol{\Lambda}_\epsilon$ shown as heatmap of probabilities.
from matplotlib.colors import LinearSegmentedColormap
from mpl_toolkits.axes_grid1 import make_axes_locatable

from lyscripts.plot.utils import COLORS

nrows, ncols = 1, 1

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(nrows=nrows, ncols=ncols))

model = shared.get_model(which="midline", load_samples=True)

fig, ax = plt.subplots(tight_layout=True)

cmap = LinearSegmentedColormap.from_list("redusz", [COLORS["gray"], COLORS["red"]])
im = ax.imshow(model.midext_evo().T, cmap=cmap)

# create an Axes on the right side of ax. The width of cax will be 5%
# of ax and the padding between cax and ax will be fixed at 0.05 inch.
divider = make_axes_locatable(ax);
cax = divider.append_axes("right", size="5%", pad=0.05);
plt.colorbar(im, cax=cax, label="Prob. of midext");

ax.set_xticks(np.arange(model.max_time + 1));
ax.set_xlabel("time-step $t$");

ax.set_yticks([0., 1.], labels=["lat.", "ext."])
```

Just as the lymphatic states, this lateralization state's evolution over the time-steps must then be marginalized over the time of diagnosis using $p(t)$.

- multiply first or second index of evolution with time-prior
- choose appropriate resulting joint, depending on patient's tumor lateralization

@@

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(nrows=nrows, ncols=ncols, aspect_ratio=1.))

@@

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(aspect_ratio=2.))

@@

>>>>

import pandas as pd
import numpy as np
import matplotlib.pyplot as plt
from matplotlib import ticker
from matplotlib.colors import LinearSegmentedColormap
import upsetplot

from lymph import models
from lyscripts import utils
from lyscripts.plot.utils import COLORS

from scripts import shared, paths, COL, CONTRA_LNLS

midline_model = shared.get_model("midline", load_samples=True)
```

```{python}
#| tags: [parameters]
#| echo: false

# width of figures. May depend on number of text columns
full = 17 # cm
half = full
# half =  8 # cm
```

@@

$$
\begin{aligned}
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \right) &= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c} \mid t \right) \\
&= \sum_{t=0}^{t_\text{max}} P(t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
\end{aligned}
$$ {#eq-bilateral-marginal}

@@

$$
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
$$ {#eq-bilateral-marginal-algebra}

@@

To account for the increased prevalence of involvement on the contralateral side when the tumor is not clearly lateralized anymore, we also model the tumor's extension over the mid-sagittal line as a binary random variable. It starts lateralized and at every time-step there is a finite probability $p_\epsilon$ that the tumor grows over the symmetry plane of the patient.

Technically, the introduction of this additional random variable doubles the space of the hidden states and therefore quadruples the size of the transition matrix $\mathbf{A}$. However, since we assume no correlation between the tumor's lateralization and the metastases in the LNLs, we can evolve the two parts separately.

The probabilities to find a patient with a clearly lateralized tumor or one that extends over the mid-sagittal line after $t$ time-steps are then given by

$$
\begin{aligned}
P(\epsilon = \texttt{False} \mid t) &= (1 - p_\epsilon)^t \\
P(\epsilon = \texttt{True} \mid t) &= 1 - (1 - p_\epsilon)^t
\end{aligned}
$$

To get the joint probability over the ipsi- and contralateral hidden states, as well as the state of the tumor's midline extension $P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right)$, we simply add the above terms to the marginalization in @eq-bilateral-marginal:

$$
P \left( \mathbf{X}^\text{i}, \mathbf{X}^\text{c}, \epsilon \right) = \sum_{t=0}^{t_\text{max}} P(t) \cdot P(\epsilon \mid t) \cdot P \left( \mathbf{X}^\text{i} \mid t \right) \cdot P \left( \mathbf{X}^\text{c} \mid t \right)
$$

Again, this can be written algebraically, by defining the vector $P(\mathbf{t}, \epsilon) = P(\mathbf{t}) \cdot P(\epsilon \mid \mathbf{t})$, to achieve the same form as in @eq-bilateral-marginal-algebra:

$$
P \left( \mathbf{X}^\text{i} = \boldsymbol{\xi}_n, \mathbf{X}^\text{c} = \boldsymbol{\xi}_m, \epsilon \right) = \left[ \boldsymbol{\Lambda}^\intercal_\text{i} \cdot \operatorname{diag} P(\mathbf{t}, \epsilon) \cdot \boldsymbol{\Lambda}_\text{c} \right]_{n,m}
$$

In @fig-midext-evo, we visualize how the prior distribution over diagnose times $P(t)$, the conditional probability of midline extension $P(\epsilon \mid t)$, and their joint $P(\epsilon, t)$ evolve over the course of a patient evolution.

```{python}
#| echo: false
#| label: fig-midext-evo
#| fig-cap: The evolution of the probability for the tumor to remain lateralized (green) or to cross the mid-sagittal plane (red).
t = np.linspace(0, 10, 11)
p_midline = {
  "lateralized": (1 - midline_model.midext_prob)**t,
  "extension": 1 - (1 - midline_model.midext_prob)**t,
}
dist = {
  "early": midline_model.get_distribution("early").pmf,
  "late": midline_model.get_distribution("late").pmf,
}
p_colors = {"lateralized": COLORS["green"], "extension": COLORS["red"]}
t_colors = {"early": COLORS["blue"], "late": COLORS["orange"]}
markers = {}

nrows, ncols = 2, 1
plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(
  nrows=nrows,
  ncols=ncols,
  width=half,
  aspect_ratio=2.5,
))

fig, axes = plt.subplots(nrows=nrows, ncols=ncols, sharex=True)
w = 0.3

for label, p in p_midline.items():
  axes[0].plot(
    t, p_midline[label], "o-",
    label=label,
    color=p_colors[label],
  )
  for i, t_stage in enumerate(["early", "late"]):
    if label == "lateralized":
      axes[0].bar(
        t + i*w - w/2, dist[t_stage],
        color=t_colors[t_stage],
        width=w,
        label=f"{t_stage} T-cat.",
      )
    axes[1].plot(
      t, p * dist[t_stage], "o-",
      c=p_colors[label],
      mfc=t_colors[t_stage],
      mec=t_colors[t_stage],
      label=f"{label}, {t_stage}",
    )

axes[0].set_ylim(0., 1.)
axes[0].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[0].set_ylabel("probability")
axes[0].grid(axis="y", color=COLORS["gray"], alpha=0.5)

axes[0].legend()

axes[1].set_xlim(min(t), max(t))
axes[1].set_xticks(t)
axes[1].set_xlabel("time-step $t$")

axes[1].set_ylim(0., 0.25)
axes[1].yaxis.set_major_formatter(ticker.StrMethodFormatter("{x:.0%}"))
axes[1].set_ylabel("probability")
axes[1].grid(axis="y", color=COLORS["gray"], alpha=0.5)

axes[1].legend()

plt.show()
```

@@

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(
  nrows=nrows,
  ncols=ncols,
  aspect_ratio=1.,
  width=full,
))

@@

plt.rcParams.update(shared.get_fontsizes())
plt.rcParams.update(shared.get_figsizes(aspect_ratio=2., width=half))

@@

## Distributions over States

```{python}
plt.rcdefaults()

mi = pd.MultiIndex.from_product(
  [[False, True]] * len(midline_model.ext.ipsi.graph.lnls),
  names=midline_model.ext.ipsi.graph.lnls.keys(),
)
series = pd.Series(midline_model.ext.ipsi.state_dist(), index=mi)
upsetplot.plot(series)
plt.show()
```
